Equations & R Code

### 2016-11-26 manuscript/coexistence-stability_ms.tex
adds PBA edits to ms; needs cleaning

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized that species can compensate for each other and stabilize
ecosystem functioning in fluctuating environments. This idea underlies
the ``insurance hypothesis'' (Yachi and Loreau 1999), which suggests
stability increases with diversity because species respond dissimilarly
to environmental conditions, broadening the range of conditions under
which the community maintains function (Loreau 2010). Diverse models all
predict a positive relationship between species richness and ecosystem
stability (Lehman and Tilman 2000, Ives and Hughes 2002, Loreau and {{de
Mazancourt}} 2013), and experimental tests tend to support such a
prediction (Tilman et al. 2006, Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to represent real-world dynamics is the subject of debate
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact that species losses in BEF experiments are not allowed to be
offset by species gains. Theoretical work on diveristy-stability
relationships typically suffers from the same limitation, ignoring the
processes by which temporal environmental variability can promote
coexistence and increase species richness (Loreau 2010, but see Chesson
et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a,
Chesson et al. 2004) and in natural communities (C{á}ceres 1997,
Descamps-Julien and Gonzalez 2005, Adler et al. 2006, Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, an increase in ecosystem
stability. However, increasing environmental variability may also
decrease ecosystem stability by increasing the fluctuations of
individual species, regardless of species richness.

The countervailing effects of environmental variability present an
interesting paradox: increasing variability should decrease ecosystem
stability, but the potential for an increase in richness might offset
the decrease in stability. Such a paradox complicates predictions about
how ecosystems will respond as environmental conditions exceed
historical ranges of variability. The unknown net effect of environment
variability may be reflected in the mixed results from empirical studies
on the diversity-stability relationship. Observational tests of the
diversity-stability relationship, which require sampling across natural
diversity gradients, have yielded positive (Hautier et al. 2014),
neutral (Valone and Hoffman 2003, Cusson et al. 2015), and negative
(Sasaki and Lauenroth 2011) relationships. In a meta-analysis of
diversity-stability relationships based on observational studies in
terrestrial ecosystems, Jiang and Pu (2009) found no significant
evidence for an effect of species richness on ecosystem stability. The
idiosyncratic results of these empirical studies contrasts with the
consistent conclusions from theoretical work.

The gap between theoretical expectations and empirical results of
richness-stability relationships reflects the divergence of theory
developed to explain species coexistence and theory developed to explain
diversity and stability. One reason these two disciplines have diverged
is because they have focused on slightly different questions.
Diversity-stability studies typically ask how ecosystem variability
responds to different levels of species richness at a given level of
environmental variability (reviewed in Kinzig et al. 2001, Loreau 2010),
whereas coexistence studies ask how the long term stability of species
coexistence responds to different levels of environmental variability
(Chesson and Warner 1981).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the direction of the diversity-stability relationship remain
  positive when species coexistence is fluctuation-dependent?
\item
  When species coexistence is fluctuation-dependent, how does increasing
  environmental variability impact ecosystem stability?
\item
  Do our answers to the previous two questions depend on the specific
  fluctuation-dependent coexistence mechanism (i.e., storage effect
  vs.~relative nonlinearity)?
\end{enumerate}

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). Regardless of the
mechanism, long-term coexistence is possible when all species can
increase when rare. In the storage effect, rare species increase by
escaping the effects of \emph{EC} covariance. This happens because
common species will experience greater than average competition
(\emph{C}) in good environment (\emph{E}) years because common species
cannot avoid intraspecific competition. Rare species do not have this
problem and can increase rapdily in a good \emph{E} year. \emph{EC}
covariance is included in our model because dormant-to-live transition
rates (\(\gamma\)) are species-specific and vary through time. In a high
\(\gamma\) year for a common species, resource uptake will be above
average, while in a high \(\gamma\) year for a rare species, resource
uptake will be below average.

Subadditive population growth buffers populations against large
population decreases in unfavorable years. It is included in our model
because we include a dormant stage with very low death rates, which
limits large population declines. In combination, subadditive population
growth limits population declines in bad \emph{E} years, and \emph{EC}
covariance ensures species can increase rapidly when rare but suffer
when common.

@@

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time because its draw down of resources starates.
Likwise, when a resource acquisative species (e.g., species A in Fig.
1C) is abundant it quickly draws down resources to levels that favor
resource conservative species. Such reciprocity ensures species can
increase when rare and stablizes coexistence (Armstrong and McGehee
1980, Chesson 2000a, Chesson et al. 2004).

@@

When we allowed the variance of the environment to determine the number
of coexisting species from an initial community of four, we found a
positive relationship between diversity and variability (Fig. 2A,C).
This was true for the storage effect, where coexistence is maintained by
fluctuating transition rates (\(\gamma\)), and for relative
nonlinearity, where coexistence is maintained by annual resource pulses.
The relationship is driven by the fact that, under both coexistence
mechanisms, the strength of species coexistence increased with
environmental variability (Fig. S1-2). Therefore, more variabile
conditions promoted species richness, creating a positive relationship
between diversity and variability.

When we performed species removals at given levels of environmental
variability, we found the typical negative diversity-variability
relationship (Fig. 2B,D). The relationship is very constrained for the
storage effect because all species have similar temporal variances.
Therefore, regardless of species identity, having more species will
always stabilize ecosystem functioning through portfolio effects. On the
contrary, the relationship is less strong for relative nonlinearity
(Fig. 2D) because each species has unique temporal variance depending on
its resource uptake curve. Depending on the specific combination of
species, two-species communities may sometimes be less variable than
three-species communities.

To analyze the above results in more detail, we simulated our model over
range of environmental variance and species pool sizes. In Fig. 3 we
show storage effect simulation results where environmental variability
determines species coexistence from a regional species pool of four
species. We also show results from nested subsets of the four species
pool (e.g., only two species in the pool instead of four) to show the
trajectory of ecosystem variability if new species are not present to
join the local community. We find that realized species richness
increases with environmental variability and, in some cases, increasing
variability can actually completely temper the effect of increasing
environmental variability. More species rich communities are less
variable on average and increase in ecosystem \emph{CV} at a slower rate
(e.g., lower slopes in log-log space; Fig. S1-3).

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Asymmetric competition
makes it more difficult for new species to enter the local community
(Fig. 3; compare top and bottom panels). Relatively high competition
also increases the rate at which ecosystem \emph{CV} increases with
environmental variance (Fig. S1-3). This is because the abundance of
inferior competitors is reduced and therefore does not influence
ecosystem \emph{CV} as much as when competition is symmetric. The
correlation of species' environmental responses also mediates the
relationship between environmental variance, species richness, and
ecosystem \emph{CV}: lower correlations make it easier for new species
to coexist and unique environmental responses are always stabilizing
(Fig. 3).

@@

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require envrionmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000a). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. S1-2).

@@

When coexistence is fluctuation-dependent, however, more interesting
things happen. By simulating communities with different species pool
sizes across a gradient of environmental variabilitym, we show that
species gains due to increasing environmental variability can temper the
direct effect of environmental variability on ecosystem variability
(Figs. 3 and 4). These results lead to two conclusions.

First, when predicting the impacts of increasing environmental
variability on ecosystem stability, the mechanism of coexistence in the
community matters. Fluctuation-dependent coexistence can buffer
ecosystems from increasing environmental variability by allowing for
species additions. Whether our theoretical predictions hold in real
communities is unknown and requires empirical tests. Doing so would
require manipulating environmental variability in communities where
coexistence is known to be fluctuation-dependent, at least in part. Such
data do exist (Angert et al. 2009), and a coupled modeling-experimental
approach could determine if our predictions hold true in real
communities.

>>>>

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi and Loreau 1999), which
suggests stability increases with diversity because species respond
dissimilarly to environmental conditions, broadening the range of
conditions under which the community maintains function (Loreau 2010). A
variety of theoretical models all predict a positive relationship
between species richness and ecosystem stability (Lehman and Tilman
2000, Ives and Hughes 2002, Loreau and {{de Mazancourt}} 2013), and
experimental tests tend to support such a prediction (Tilman et al.
2006, Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact BEF experimental protocols do not allow species gains to offset
species losses. Theoretical work on diversity-stability relationships
typically suffers from the same limitation: it recognizes the role of
environmental variability in driving population fluctuations which
destabilize ecosystems, but ignores the potential for environmental
variability to promote species richness and thereby help stabilize
ecosystems (Loreau 2010, but see Chesson et al. 2001). THE CHANGES I
MADE IN THE PREVIOUS PARAGRAPH ARE REDUNDANT WITH THE END OF THE NEXT
PARAGRAPH. YOU MIGHT NEED TO REORGANIZE, BUT I THOUGHT IT WAS IMPORTANT
TO GET TO THESE POINTS SOONER THAN LATER.

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a,
Chesson et al. 2004) and in natural communities (C{á}ceres 1997,
Descamps-Julien and Gonzalez 2005, Adler et al. 2006, Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, an increase in ecosystem
stability. However, increasing environmental variability may also
decrease ecosystem stability by increasing the fluctuations of
individual species, regardless of species richness. These countervailing
effects of environmental variability present an interesting paradox:
increasing variability should decrease ecosystem stability, but the
potential for an increase in richness might offset the decrease in
stability. Such a paradox complicates predictions about how ecosystems
will respond to predicted departures from the historical ranges of
environmental variability.

The opposing effects of environmental variability on stability might
explain the mixed results from empirical studies on the
diversity-stability relationship. Observational tests of the
diversity-stability relationship, which require sampling across natural
diversity gradients, have yielded positive (Hautier et al. 2014),
neutral (Valone and Hoffman 2003, Cusson et al. 2015), and negative
(Sasaki and Lauenroth 2011) relationships. In a meta-analysis of
diversity-stability relationships based on observational studies in
terrestrial ecosystems, Jiang and Pu (2009) found no significant
evidence for an effect of species richness on ecosystem stability. The
idiosyncratic results of these empirical studies contrast with the
consistent conclusions from theoretical work which ignores feedbacks
between variability and richness.

The gap between theoretical expectations and empirical results of
richness-stability relationships reflects the divergence of theory
developed to explain species coexistence and theory developed to explain
diversity and stability. One reason these two disciplines have diverged
is because they have focused on different questions. Diversity-stability
studies typically ask how ecosystem variability responds to different
levels of species richness at a given level of environmental variability
(reviewed in Kinzig et al. 2001, Loreau 2010), whereas coexistence
studies ask how species richness responds to different levels of
environmental variability (Chesson and Warner 1981).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the diversity-stability relationship remain positive when species
  coexistence is fluctuation-dependent?
\item
  When species coexistence is fluctuation-dependent, how does increasing
  environmental variability impact ecosystem stability?
\item
  Do our answers to the previous two questions depend on the specific
  fluctuation-dependent coexistence mechanism (i.e., storage effect
  vs.~relative nonlinearity)?
\end{enumerate}

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). Regardless of the
mechanism, long-term coexistence is possible when all species can
increase when rare. In the storage effect, rare species increase by
escaping the effects of \emph{EC} covariance. This happens because
common species will experience greater than average competition
(\emph{C}) in good environment (\emph{E}) years because common species
cannot avoid intraspecific competition. Rare species do not have this
problem and can increase rapidly in a good \emph{E} year. \emph{EC}
covariance is included in our model because dormant-to-live transition
rates (\(\gamma\)) are species-specific and vary through time. In a high
\(\gamma\) year for a common species, resource uptake will be above
average, while in a high \(\gamma\) year for a rare species, resource
uptake will be below average.

Subadditive population growth buffers populations against large
population decreases in unfavorable years. It is included in our model
through a dormant stage with very low death rates, which limits large
population declines. In combination, subadditive population growth
limits population declines in bad \emph{E} years, and \emph{EC}
covariance ensures species can increase rapidly when rare but suffer
when common.

@@

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time because its draw down of resources starates.
Likwise, when a resource acquisitive species (e.g., species A in Fig.
1C) is abundant, it quickly draws down resources to levels that favor
resource conservative species. Such reciprocity ensures species can
increase when rare and stablizes coexistence (Armstrong and McGehee
1980, Chesson 2000a, Chesson et al. 2004).

@@

When we allowed the variance of the environment to determine which of
four initial species coexist, we found a positive relationship between
richness and variability (Fig. 2A,C). This was true for the storage
effect, where coexistence is maintained by fluctuating transition rates
(\(\gamma\)), and for relative nonlinearity, where coexistence is
maintained by annual resource pulses. The relationship is driven by the
fact that, under both coexistence mechanisms, the strength of species
coexistence increased with environmental variability (Fig. S1-2).
Therefore, more variable conditions promoted species richness, creating
a positive relationship between diversity and variability.

When we performed species removals but held environmental variability at
given levels, we found the typical negative diversity-variability
relationship (Fig. 2B,D). The relationship was very constrained for the
storage effect because all species have similar temporal variances.
Therefore, regardless of species identity, having more species will
always stabilize ecosystem functioning through portfolio effects. On the
contrary, the relationship was weaker for relative nonlinearity (Fig.
2D) because each species has unique temporal variance depending on its
resource uptake curve. Depending on the specific combination of species,
two-species communities were sometimes less variable than three-species
communities.

To analyze the above results in more detail {[}THIS IS VAGUE, WHAT WAS
THE OBJECTIVE?{]}, we simulated our model over a range of environmental
variance and species pool sizes. In Fig. 3 we show storage effect
simulation results where environmental variability determines species
coexistence from a regional species pool of four species. We also show
results from nested subsets of the four species pool (e.g., only two
species in the pool instead of four) to show the trajectory of ecosystem
variability if new species are not present to join the local community.
I FIND THE PREVIOUS TWO SENTENCES CONFUSING and METHODS-Y. CUT? Realized
species richness increased with environmental variability and, in some
cases, increasing variability completely offset the effect of increasing
environmental variability on stability. More species rich communities
were less variable on average and increased in ecosystem \emph{CV} at a
slower rate than communities with fewer species (e.g., lower slopes in
log-log space; Fig. S1-3).

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Asymmetric competition
makes it more difficult for new species to enter the local community
(Fig. 3; compare top and bottom panels). Relatively high competition
also increases the rate at which ecosystem \emph{CV} increases with
environmental variance (Fig. S1-3) because the abundance of inferior
competitors is reduced and therefore does not influence ecosystem
\emph{CV} as much as when competition is symmetric. The correlation of
species' environmental responses also mediates the relationship between
environmental variance, species richness, and ecosystem \emph{CV}: lower
correlations make it easier for new species to enter the community and
unique environmental responses are always stabilizing (Fig. 3).

@@

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require envrionmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000a). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. S1-2), but the increase in
environmental variability also increases ecosystem variability.

@@

When coexistence is fluctuation-dependent, however, the outcome is less
certain. By simulating communities with different species pool sizes
across a gradient of environmental variabilitym, we showed that species
gains due to increasing environmental variability can buffer the direct
effect of environmental variability on ecosystem variability (Figs. 3
and 4). These results lead to two conclusions.

First, when predicting the impacts of increasing environmental
variability on ecosystem stability, the mechanism of coexistence in the
community matters. Fluctuation-dependent coexistence can buffer
ecosystems from increasing environmental variability by promoting
increased species richness. Whether our theoretical predictions hold in
real communities is unknown and requires empirical tests. Doing so would
require manipulating environmental variability in communities where
coexistence is known to be fluctuation-dependent, at least in part. Such
data do exist (Angert et al. 2009), and a coupled modeling-experimental
approach could determine if our predictions hold true in real
communities.
